---
title: "Project 1_RR"
author: "Goga"
date: "2023-05-23"
output: html_document
---
##Assignment Instructions
1.Code for reading in the dataset and/or processing the data
2.Histogram of the total number of steps taken each day
3.Mean and median number of steps taken each day
4.Time series plot of the average number of steps taken
5.The 5-minute interval that, on average, contains the maximum number of steps
6.Code to describe and show a strategy for imputing missing data
7.Histogram of the total number of steps taken each day after missing values are imputed
8.Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9.All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

##Step 1
##Code for reading in the dataset and/or processing the data
```{r, echo = TRUE}
setwd("C:/Users/gogab/Documents/R Studio files/Reproductible research")
activity <- read.csv("C:/Users/gogab/Documents/R Studio files/Reproductible research/repdata_data_activity/activity.csv")
```

Exploring the basics of this dataset "activity.csv"
```{r}
dim(activity)
names(activity)
head(activity)
str(activity)
#number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
#transforming date format using lubridate
library(lubridate)
activity$date <- ymd(activity$date)
length(unique(activity$date))
```
##Step 2
##Histogram of the total number of steps taken each day
##Step 3
##Mean and median number of steps taken each day
```{r, echo = TRUE}
library(ggplot2)
total.steps <- tapply(activity$steps, activity$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth = 1000, xlab = "total number of steps taken each day")
dev.off()
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

##Step 4
##Time series plot of the average number of steps taken
```{r, echo = TRUE}
library(ggplot2)
averages <- aggregate(x = list(steps = activity$steps), by = list(interval = activity$interval),
                      FUN = mean, na.rm = TRUE)
ggplot(data = averages, aes(x = interval, y = steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
dev.off()
```

##Step 5
##The 5-minute interval that, on average, contains the maximum number of step
```{r, echo = TRUE}
averages[which.max(averages$steps),]
```


##Step 6
##Code to describe and show a strategy for imputing missing data.
For the purpose I will use the mean/mode value substitution strategy to impute missing values. By using the mean values we'll substitute out the missing values in the original data set
First, it is useful to know what are the distributions of missing values by date and interval.
```{r how_many_missing}
missing <- is.na(activity$steps)
# How many missing
table(missing)
```

All of the missing values are filled in with mean value for that 5-minute
interval.

```{r}
# Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- activity
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```
## Step 7.
## Histogram of the total number of steps taken each day after missing values are imputed

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
dev.off()
```

Mean and median values are higher after imputing missing data. The reason is that in the original data, there are some days with `steps` values `NA` for any `interval`. The total number of steps taken in such days are set to 0s by default. 

## Step 8
## Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```
Plots of average number of steps taken on weekdays and weekends.
```{r}
averages <- aggregate(steps ~ interval + day, data = filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
dev.off()
```
